Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-30 paper_nips/paper.tex
possible rephrasing
rest of intro

To test, on the other hand,
the neural correlates underlying mental tasks,
the general linear model (GLM) is the dominant approach \cite{friston94}.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI)
elucidate the influence of one brain region on another conditioned
by experimental tasks \cite{friston97}.
% XXX Michael more and more or ever more:
As a last example, \suggestremove{always} \suggestadd{more and} more 
neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to task-associated (i.e., \textit{labeled})
data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses typically yield
results in a voxel space.
This ignores the fact that the BOLD
signal exhibits spatially distributed activity patterns.
%
Second, existing supervised neuroimaging analyses
do not acknowledge the abundance
of easily acquired resting-state data \cite{biswaldiscovery}.
These may allow better sampling the manifold of brain states
due to the high task-rest similarities of neural activity patterns,
as observed using ICA \cite{smith2009}.

Both these biological properties can be conjointly exploited in a mixed
 (i.e., use task and rest data)
low-rank (i.e., perform network decomposition)
approach.
%
The integration of brain-network discovery into
supervised classification formally yields a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Architecture.}
  Linear autoencoders find an optimal compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  is thus guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders
are one-layered learning models that condense the input data to
local and global representations
by improving reconstruction from them.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex transformation \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.

In the present investigation,
a linear autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification in an identical statistical learning
problem.
%
From the perspective of dictionary learning, the first layer represents
the discovered set of basis functions
whose linear combinations are learned
by the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
put probability mass on the most neurobiologically
valid models in the hypothesis set.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

>>>>

To test, on the other hand,
the neural correlates underlying mental tasks,
the general linear model (GLM) is the dominant approach \cite{friston94}.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI)
elucidate the influence of one brain region on another conditioned
by experimental tasks \cite{friston97}.
% XXX Michael more and more or ever more:
As a last example, \suggestremove{always} \suggestadd{more and} more 
neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to task-associated (i.e., \textit{labeled})
data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses typically yield
results in a voxel space.
This ignores the fact that the BOLD
signal exhibits spatially distributed \suggestremove{activity} patterns
\suggestadd{of coherent activity}.
%
Second, existing supervised neuroimaging analyses
\suggestremove{do not acknowledge} \suggestadd{cannot exploit} the abundance
of easily acquired resting-state data \cite{biswaldiscovery}.
These may allow better \suggestremove{sampling} \suggestadd{discovery of} 
the manifold of brain states
due to the high task-rest similarities of neural activity patterns,
as observed using ICA \cite{smith2009}.

% XXX Michael: This parenthesis structure should probably be unravelled into
% XXX a sentence.
Both these biological properties can be conjointly exploited in a mixed
 (i.e., using task and rest data)
low-rank (i.e., performing network decomposition)
approach.
%
The integration of brain-network discovery into
supervised classification formally yields a semi-supervised learning
% XXX Michael: No, it doesn't necessarily. You can do that with labeled data
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Architecture.}
  Linear autoencoders find an optimal compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  is thus guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders
are \suggestremove{one-layered} learning models 
that condense the input data to
\suggestremove{local and global} \suggestadd{(typically lower-dimensional)} 
representations
by \suggestremove{improving} \suggestadd{optimizing}
reconstruction from them.
%
% XXX Michael: In fact, they behave like a non-centered PCA, 
% XXX which amounts to just doing an SVD
% XXX which btw is NOT a convex optimization problem, although easy to 
% XXX calculate and with a global optimum.
% XXX Autoencoders are also not one-layered by default. Let's make sure that
% XXX we don't trip up by getting the technicalities wrong and getting 
% XXX reviewers annoyed at us because of it.
They behave like a \suggestadd{(truncated, non-centered)} PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a \suggestadd{non-}convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex transformation \cite{le2011ica}.
% XXX What convex transformation? Why is that relevant? Convexity is only
% XXX relevant for the loss
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Hence, autoencoders can flexibly project the neuroimaging data
onto the main axes of variation and thus % XXX Michael <- that's PCA
reverse-engineer properties of the underlying
neural processes \cite{olshausen96}.
% XXX Michael ^ I don't get that last part nor the citation.

In the present investigation,
a linear autoencoder will be \suggestremove{fed by} \suggestadd{fit to} 
(unlabeled) rest data and
integrated as \suggestadd{a rank-reducing} bottleneck
into a \suggestremove{low-rank} \suggestadd{multinomial} 
logistic regression \suggestremove{fed by} \suggestadd{fit to} 
(labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification in an identical statistical learning
problem.
%
From the perspective of dictionary learning, the first layer represents
% XXX Michael: This is the crucial difference: The first layer is an analysis
% XXX operation: it extracts information in some way. It may be extracting
% XXX activations for basis functions, in which case it would be the pseudo-
% XXX inverse of those basis functions. But unless they are orthogonal, these
% XXX weights definitely don't represent basis functions themselves. It is 
% XXX the second layer of the autoencoder that represents basis functions. If
% XXX weights are tied, they are one and the same thing, with interesting
% XXX implications. But in general, only the second layer represents a basis.
\suggestadd{projectors to} the discovered set of basis functions
whose linear combinations are learned
by the second layer \cite{olshausen96}.
% XXX Michael: ^ This needs to be changed around: This first layer obtains
% XXX activations, the second layer weights are basis functions which are
% XXX multiplied to these activations
%
Neurobiologically, this allows 
delineating a low-dimensional \suggestremove{manifold}
\suggestadd{vector space} of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
put probability mass on the most neurobiologically
% XXX Michael: ^ MEH, we are not really in a probabilistic setting ...
valid models in the hypothesis set.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
% XXX Michael: Taken together, the important modes of variation in brain dynamics and the neural correlates subserving mental operations have mostly been studied in isolation? Are they taken together, or isolated? :)
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.
